---
title: "HW 4"
author: "Emma Johnson"
date: "February 19, 2025"
output:
  pdf_document: default
  html_document: default
subtitle: STAT 5211
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR2)
library(tidyverse)
library(MASS)
```

# 6

## a)

Using the **logistic function**,

```{r 6a}
# p(x) =
exp(-6 + 0.05*40 + 1*3.5) / (1 + exp(-6 + 0.05*40 + 1*3.5))
```

The probability of a student who studies for 40 hours and has an undergrad GPA of 3.5 getting an A in the class is approximately 0.3775 under the logistic model.

## b)

Solving the **logit** function for $X_1$,

```{r 6b}
# X1 =
(log(0.5 / (1 - 0.5)) + 6 - 1*3.5) / 0.05
```

The student in part (a) would need to study for 50 hours to have a 50% chance of getting an A in the class.

# 9

## a)

Consider the odds $\frac{p(x)}{1 - p(x)} = 0.37$. Rearrange to solve for $p(x)$, that is,

```{r 9a}
# p(x) =
.37 / 1.37
```

Approximately 27.01% of people with an odds of 0.37 of defaulting on their credit card payment will in fact default.

## b)

Plugging in $p(x) = 0.16$ into the odds,

```{r 9b}
# odds = 
0.16 / (1 - 0.16)
```

The odds that an individual with a 16% chance of defaulting on her credit card payment *actually* defaulting are approximately 0.19.

# 13

```{r 13}
weekly <- ISLR2::Weekly
```

## a)

```{r 13a}
summary(weekly)

pairs(weekly)

cor(weekly[-9])

attach(weekly)
plot(Volume)
```

When observing the **pairs** plot, there appears to be a logistic relationship between *Direction* and *Today*.

There appears to be little correlation between any variables, aside from a strong positive correlation between *Year* and *Volume*. In line with the **Smarket** data, there is little correlation between the lag variables and today's returns. Further, *Volume* increases over time (indexed chronologically).

## b)

```{r 13b}
attach(weekly)

mod1 <- glm(
  Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
  family = binomial
  )
summary(mod1)
```

The only statistically significant variable within this model is *Lag2* at the 0.05 significance level. The *Intercept* is the only other term yielding significance within the summary of the model, at a 0.01 significance level.

## c) 

```{r 13c}
glm.probs <- predict(mod1, type = "response")
glm.pred <- rep("Down", 1089)
glm.pred[glm.probs > .5] = "Up"

# Confusion Matrix:
table(glm.pred, Direction)

# Correctly classified:
mean(glm.pred == Direction)
```

Under the confusion matrix, which helps identify the class of the observations in accordance to the categorical response, *Direction*, it appears that the first model only classifies the correct *Direction* as Up or Down about 56.11% of the time. This means nearly half of the observations are incorrectly categorized, making the model slightly better than a 50/50 guessing game.

Since the model was **trained** and **tested** with the same set of observations, the **training error rate** is about 43.89%, which typically underestimates the true **test error rate**. To assess the accuracy of the model, a smaller portion of the data should be tested and the rest should be **held out** to examine a more realistic error rate once predictions are carried out. 

## d) 

```{r 13d}
train <- (Year < 2009) # training data from 1990 - 2008
weekly.2009 <- weekly[!train, ] # held out data from 2009 - 2010
Direction.2009 <- Direction[!train]

mod2 <- glm(
    Direction ~ Lag2,
    data = weekly, 
    family = binomial, 
    subset = train
  )
glm.probs <- predict(mod2, weekly.2009,
    type = "response")

glm.pred <- rep("Down", 104)
glm.pred[glm.probs > .5] <- "Up"
table(glm.pred, Direction.2009)

mean(glm.pred == Direction.2009)
```

This time, using a training model from the set of observations from 1990 to 2008 and performing predictions on a held-out data set from 2009 to 2010, performs much better. This new model, using only the significant *Lag2* predictor variable, classifies the correct *Direction* around 62.5% of the time, which is surely a step up from 50/50 guessing. Further, the **training error rate** is 37.5%, which has dropped from the first model attempt of 43.89%.

# 14

## a)

```{r 14a}
auto <- ISLR2::Auto
auto$mpg01 <- ifelse(auto$mpg > median(auto$mpg), 1, 0)
```

## b)

```{r 14b pairs}
pairs(auto)
```

Aside from *mpg*, of course, the variables that may perform well in predicting *mpg01* are *horsepower*, *weight*, and *acceleration*.

```{r 14b scatterplots}
attach(auto)
par(mfrow = c(2, 2))
plot(horsepower, mpg01)
plot(weight, mpg01)
plot(acceleration, mpg01)
```

```{r boxplots}
par(mfrow = c(1, 3))
boxplot(horsepower ~ mpg01, 
        horizontal = T,
        col = c("lightblue", "lightgreen"))
boxplot(weight ~ mpg01,
        horizontal = T, 
        col = c("lightblue", "lightgreen"))
boxplot(acceleration ~ mpg01,
        horizontal = T, 
        col = c("lightblue", "lightgreen"))
```

As seen through the boxplots, it appears that the variables horsepower, weight, and acceleration each generally correspond to distinct values for *mpg01*. Note that if *mpg01* was plotted against a weakly correlated variable, such as *cylinders*, the error bars would likely overlap more, especially within the IQR.


```{r mer, include = FALSE}
library(patchwork)
box <- function(var) {
ggplot(auto, aes(x = {{var}}, y = as.factor(mpg01),
                 fill = as.factor(mpg01))) + 
  ylab("mpg01") +
  xlab(deparse(substitute(var))) +
  geom_boxplot() + 
  theme_classic() +
  theme(legend.position = "none")
}

box(horsepower) +  box(weight) + box(acceleration)
```


## c)

```{r 14c}
set.seed(5)  # reproducibility

# Sample to split 80% train & 20% test 
train_ind <- sample(1:nrow(auto), size = 0.8 * nrow(Auto))
train <- rep(FALSE, nrow(auto))
train[train_ind] <- TRUE  # training set: 313 observations

test <- auto[!train, ]  # held out data (testing): 79 observations
mpg01.test <- mpg01[!train]  # testing mpg01 values
```

## f)

```{r 14f}
# Selecting variables
auto <- auto[, !names(auto) %in% c("mpg", "name")]
mod1 <- glm(
  mpg01 ~ 1,
  data = auto, 
  family = binomial, 
  subset = train
  )
mod2 <- glm(
  mpg01 ~ .,
  data = auto, 
  family = binomial, 
  subset = train
  )
stepAIC(mod1, direction = "both", 
        scope = list(lower=mod1, upper=mod2),
        trace = F) # AIC selection

# Final training model
mod.train <- glm(
  mpg01 ~ displacement + weight + year + horsepower,
  data = auto, 
  family = binomial, 
  subset = train
  )
summary(mod.train)

# Predicting test set observations for mpg01
glm.probs <- predict(mod.train, test,
    type = "response")

glm.pred <- rep(0, nrow(test))
glm.pred[glm.probs > .5] <- 1
table(glm.pred, mpg01.test) # confusion matrix

mean(glm.pred == mpg01.test) # training accuracy rate
mean(glm.pred != mpg01.test) # training error rate
```

The **test error** of the model predicting *mpg01* using *displacement*, *weight*, *year*, and *origin* is approximately 8.86%, which seems pretty outstanding. 