---
title: "HW 5"
author: "Emma Johnson"
date: "February 28, 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR2)
library(tidyverse)
library(MASS)
```

# 13

```{r 13}
weekly <- ISLR2::Weekly
```

## e)

**LDA**

```{r 13e}
attach(weekly)
train <- (Year < 2009) # training data from 1990 - 2008
weekly.2009 <- weekly[!train, ] # held out data from 2009 - 2010
Direction.2009 <- Direction[!train]

# LDA Model:
lda.fit <- lda(
  Direction ~ Lag2, 
  data = weekly,
  subset = train
  )
lda.fit

# Predictions using LDA model:
lda.class <- predict(lda.fit, weekly.2009)$class

# Confusion matrix:
table(lda.class, Direction.2009)

# Fraction of correct predictions:
mean(lda.class == Direction.2009) # same as logistic

# Test error:
mean(lda.class != Direction.2009)
```

The LDA model of the *Weekly* data set predicts the correct *Direction* based on *Lag2* (for data in 2009 and 2010) 62.5% of the time. This is equivalent to the results from the logistic regression model. 

## f)

**QDA**

```{r 13f}
# QDA Model:
qda.fit <- qda(
  Direction ~ Lag2, 
  data = weekly,
  subset = train
  )
qda.fit

# Predictions using QDA model:
qda.class <- predict(qda.fit, weekly.2009)$class

# Confusion matrix:
table(qda.class, Direction.2009)

# Fraction of correct predictions:
mean(qda.class == Direction.2009)

# Test error:
mean(qda.class != Direction.2009)
```

Under the QDA model, *Direction* is predicted correctly approximately 58.65% of the time.

## g)

**KNN** with k = 1

```{r 13g}
library(class)
set.seed(5)

# Updating training & testing sets:
train.X <- matrix(Lag2[train], ncol = 1)
test.X <- matrix(Lag2[!train], ncol = 1)
train.class <- Direction[train]
test.class <- Direction[!train]

# Predictions using KNN:
knn.pred <- knn(train.X, test.X, train.class, k = 1)
mean(knn.pred != "Up") # 50% of observations classified "Up" and 50% "Down"

# Confusion Matrix:
table(knn.pred, test.class)

# Fraction of correct predictions:
mean(knn.pred == test.class)

# Test Error:
mean(knn.pred != test.class)
```

Under KNN with $k=1$, *Direction* is predicted correctly 50% of the time, not any better than simply guessing.

## h)

**Naive Bayes**

```{r 13h}
library(e1071)

# Naive Bayes Model:
nb.fit <- naiveBayes(
  Direction ~ Lag2,
  data = weekly,
  subset = train
  )
nb.fit

# Predictions using Naive Bayes:
nb.class <- predict(nb.fit, weekly.2009)

# Confusion matrix:
table(nb.class, Direction.2009)

# Fraction of correct predictions:
mean(nb.class == Direction.2009)

# Test error:
mean(nb.class != Direction.2009)
```

Under naive Bayes, *Direction* is predicted correctly approximately 58.65% of the time, similar to the QDA model.

## i)

The method that provides the best results on the *weekly* data is either the logistic regression model or the LDA model, since their error rate and accuracy rates are equivalent.

# 14

```{r 14}
auto <- ISLR2::Auto
auto$mpg01 <- ifelse(auto$mpg > median(auto$mpg), 1, 0)
```

## c)

```{r 14c}
attach(auto)
set.seed(5)  # reproducibility

# Sample to split 80% train & 20% test 
train_ind <- sample(1:nrow(auto), size = 0.8 * nrow(auto))
train <- rep(FALSE, nrow(auto))
train[train_ind] <- TRUE  # training set: 313 observations

test <- auto[!train, ]  # held out data (testing): 79 observations
mpg01.test <- mpg01[!train]  # test set mpg01 (true) values 
```

## e)

**LDA**

```{r 14e}
# LDA Model:
lda.fit <- lda(
  mpg01 ~ displacement + weight + year + horsepower,
  data = auto,
  subset = train
  )
lda.fit

# Predictions using LDA model:
lda.class <- predict(lda.fit, test)$class

# Confusion matrix:
table(lda.class, mpg01.test)

# Fraction of correct predictions:
mean(lda.class == mpg01.test) # training accuracy rate

# Test error:
mean(lda.class != mpg01.test)
```

The LDA model of the *Auto* data set, predicts *mpg01* (i.e. above or below the median mpg) based on *displacement*, *weight*, *year*, and *horsepower*, with a **test error** of approximately 7.59%. This is better than the results from the logistic regression model. 

## f)

**QDA**

```{r 14f}
# QDA Model:
qda.fit <- qda(
  mpg01 ~ displacement + weight + year + horsepower,
  data = auto,
  subset = train
  )
qda.fit

# Predictions using QDA model:
qda.class <- predict(qda.fit, test)$class

# Confusion matrix:
table(qda.class, mpg01.test)

# Fraction of correct predictions:
mean(qda.class == mpg01.test) # training accuracy rate

# Test error:
mean(qda.class != mpg01.test)
```

Under the QDA model, the **test error** is approximately 7.59%, similar to the results of the LDA model.

## g)

**Naive Bayes**

```{r 14g}
# Naive Bayes Model:
nb.fit <- naiveBayes(
  mpg01 ~ displacement + weight + year + horsepower,
  data = auto,
  subset = train
  )
nb.fit

# Predictions using Naive Bayes:
nb.class <- predict(nb.fit, test)

# Confusion matrix:
table(nb.class, mpg01.test)

# Fraction of correct predictions:
mean(nb.class == mpg01.test)

# Test error:
mean(nb.class != mpg01.test)
```

Under naive Bayes, the **test error** is approximately 8.86%.

## h)

**KNN**

```{r 14h}
# Standardizing the quantitative variables:
standardized.X  <- scale(auto[,c(3:5, 7)]) # displacement, horsepower, weight, year
apply(standardized.X, 2, mean) # means approx 0
apply(standardized.X, 2, sd) # st. devs = 1

# Updating training & testing sets:
train.X <- standardized.X[train, ]
test.X <- standardized.X[!train, ]
train.class <- mpg01[train]
test.class <- mpg01[!train]
```

Testing **k = 1**:

```{r 14h k of 1}
set.seed(5)

# Predictions using KNN with k = 1:
knn.pred <- knn(train.X, test.X, train.class, k = 1)
mean(knn.pred == 1) # 60.76% observations classified as a 1 (i.e. above median mpg)

# Confusion Matrix:
table(knn.pred, test.class)

# Fraction of correct predictions:
mean(knn.pred == test.class)

# Test Error:
mean(knn.pred != test.class)
```

Testing **k = 2**:

```{r 14h k of 2}
set.seed(5)

# Predictions using KNN with k = 2:
knn.pred <- knn(train.X, test.X, train.class, k = 2)
mean(knn.pred == 1) # 54.43% observations classified as a 1 (i.e. above median mpg)

# Confusion Matrix:
table(knn.pred, test.class)

# Fraction of correct predictions:
mean(knn.pred == test.class)

# Test Error:
mean(knn.pred != test.class)
```

Testing **k = 3**:

```{r 14h k of 3}
set.seed(5)

# Predictions using KNN with k = 3:
knn.pred <- knn(train.X, test.X, train.class, k = 3)
mean(knn.pred == 1) # 56.96% observations classified as a 1 (i.e. above median mpg)

# Confusion Matrix:
table(knn.pred, test.class)

# Fraction of correct predictions:
mean(knn.pred == test.class)

# Test Error:
mean(knn.pred != test.class)
```

Testing **k = 10**:

```{r 14h k of 10}
set.seed(5)

# Predictions using KNN with k = 10:
knn.pred <- knn(train.X, test.X, train.class, k = 10)
mean(knn.pred == 1) # 59.49% observations classified 1 (i.e. above median mpg)

# Confusion Matrix:
table(knn.pred, test.class)

# Fraction of correct predictions:
mean(knn.pred == test.class)

# Test Error:
mean(knn.pred != test.class)
```

Under KNN, the **test error** is minimized when $k=1$ at approximately 3.80%. Notably, the other trial with $k=2,3,10$ each resulted in the same, if not lower, **test error** rate than the previous three methods, respectively.

Therefore, it appears that KNN is the best method for accurately predicting *mpg01*.
